### 2010-03-09 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: Section 'type theory' done. Good night.

@@

\begin{structure}
<- Presentation of the formalism
    <- Standard presentation
        -> No novelty here
    <- 3 judgments [equation]
        -> Context validity
        -> Typing judgements
        -> Equality judgements
\end{structure}

\[
\begin{array}{ll}
\Gamma\vdash\Valid & \mbox{\(\Gamma\) is a valid context, giving types to variables} \\
\Gamma\vdash \Bhab{t}{T} & \mbox{term \(t\) has type \(T\) in context \(\Gamma\)} \\
\Gamma\vdash \Bhab{s \equiv t}{T} & \mbox{\(s\) and \(t\) are equal at type \(T\) in context \(\Gamma\)} \\
\end{array}
\]

\begin{structure}
    <- Invariants [equation]
        -> By induction on derivations
\end{structure}

The systems of inference rules will be formulated to ensure that that the
following implications always hold by induction on derivations.
\[
\begin{array}{l@{\;\;\Rightarrow\;\;}l}
\Gamma\vdash \Bhab{t}{T}            & \Gamma\vdash\Valid \;\wedge\; \Gamma\vdash\Type{T} \\
\Gamma\vdash s \equiv \Bhab{t}{T}   & \Gamma\vdash \Bhab{s}{T} \;\wedge\; \Gamma\vdash \Bhab{t}{T} \\
\Gamma;\xS;\Delta \vdash J          & \Gamma\vdash \Bhab{s}{S} \;\Rightarrow\; 
                                           \Gamma;\Delta[s/x] \vdash J[s/x] \\
\end{array}
\]

\begin{structure}
    <- Judgemental equality
        <- Presentation independant of particular implementation choice
        -> Model in Agda, intensional
        -> Used in Epigram, OTT
\end{structure}

\begin{structure}
<- Context validity [figure]
    <- Not much to be said
\end{structure}

@@

\begin{structure}
<- Typing judgements [figure]
    <- Set in Set
        -> For simplicity of presentation
        -> Assume that a valid stratification can be inferred
            <- Harper-Pollack, Luo, Courant
        -> See later discussion
    <- Standard presentation of Pi and Sigma types
\end{structure}

@@

\begin{structure}
<- Judgemental equality [figure]
    <- symmetry, reflexivity, and transitivity
    <- beta-rules for lambda and pair
    <- xi-rule for functions
    -> Agnostic in the notion of equality
        <- Doesn't rely on a ``propositional'' equality
\end{structure}

@@

\begin{structure}
!!! Need Help !!!
<- Meta-theoretical properties
    <- Assuming a stratified discipline
    <> The point here is to reassert that dependent types are not evil, 
       there is no non-terminating type-checker, or such horrible lies <>
    -> Strongly normalizing
        -> Every program terminates
    -> Type-checking terminates
    ???
\end{structure}

@@

\begin{structure}
<- Motivation
    <- Finite sets could be encoded with Unit and Bool
        /> Hinder the ability to name things
    -> For convenience
        <- Named elements
        <- Referring by name instead of code
        -> Types as coding presentation
            /> Also as coding representation!
\end{structure}

\begin{structure}
<- Implementation [figure]      
    <- Tags
        -> Purely informational token
    <- EnumU
        -> Universe of finite sets
    <- EnumT e
        -> Elements of finite set e
\end{structure}

@@

\begin{structure}
<- Equipment
    <- \spi operator
        <- Equivalent of Pi on finite sets
        <- First argument: (finite) domain
        <- Second argument: for each element of the domain, a co-domain
        -> Inhabitant of \spi: right-nested tuple of solutions
            <- Skip code for space reasons
    <- switch operator
        <- case analyses over x
        <- index into the \spi tuple to retrieve the corresponding result
\end{structure}

\[\stkl{
%% spi
\F{$\pi$} : \PITEL{\V{e}}{\EnumU}
            \PITEL{\V{P}}{\EnumT{e} \To \Set} \To \Set 
\\
%% switch
\begin{array}{@{}ll}
\F{switch} : & \PITEL{\V{e}}{\EnumU}
               \PITEL{\V{P}}{\EnumT{e} \To \Set} \\
             & \PITEL{\V{b}}{\spi{e}{P}}
               \PITEL{\V{x}}{\EnumT{e}} \To P~x
\end{array}
}\]

\caption{Operators on finite sets}
\label{fig:operator-finite-set}

\begin{structure}
<- Equivalent to having a function space over finite sets
    /> Made non-obvious by low-level encodings
        <- General issue with codes
             -> Need to provide an attractive presentation to the user
    -> Types seem to obfuscate our reading
        <- Provide ``too much'' information
        /> False impression: information is actually waiting to be used more widely
        -> See next Section
\end{structure}

@@

\begin{structure}
<- Bidirectional type-checking [ref. Turner,Pierce]
    -> Separating type-checking from type synthesis
    <- Type checking: push terms into types
        <- Example: |Pi S T :>: \ x . t| allows us to drop annotation on lambda
    <- Type inference: pull types out of terms
        <- Example: |x : S l- x :<: S| gives us the type of x
\end{structure}

\begin{structure}
<- Formalization: type propagation
    <- Motivation
        -> High-level syntax
            -> exprIn: types are pushed in
                <- Subject to type *checking*
            -> exprEx: types are pulled from
                <- Subject to type *synthesis*
        -> Translated into our low-level type theory
        -> Presented as judgements
    -> Presentation mirors typing rule of [figure] 
        -> Ignore identitical judgements
\end{structure}

\begin{structure}
<- Type checking [figure]
    <- Push a type in an exprIn
    <- Result in a full term
    -> *Use* the type to build the term!
        -> Domain and co-domain propagation for Pi and Sigma
        -> Translation of 'tags into EnumTs
        -> Translation of ['tags ...] into EnumUs
        -> Finite function space into switch
\end{structure}

\input{figure_type_checking}

\caption{Type checking}
\label{fig:type-checking}

\begin{structure}
<- Type synthesis [figure]
    <- Pull a type out of an exprEx
    <- Result in a full term, together with its type
    -> Do *not* need to specify types
        -> Extracting a term from the context
        -> Function application
        -> Projections
\end{structure}

\input{figure_type_synthesis}

\caption{Type synthesis}
\label{fig:type-synthesis}

>>>>

Our work is deeply rooted in the realm of dependently-typed functional
languages. In particular, we have developped this technology in our
on-going work on Epigram~\cite{pigs:epigram}. However, in this paper,
we strive to present our technology in the broadest setting
possible. To this end, we are going to present a type theory from the
ground-up.

To the reader new to type theory, this first section will provide a
gentle introduction to dependently-typed systems. To the reader
familiar with type theory, she will recognize the basis of her
favorite type theory, be it Agda or Coq for instance. This scene sets,
we have a common ground for discussion.

@@

\begin{wstructure}
<- Presentation of the formalism
    <- Standard presentation
        -> No novelty here
    <- 3 judgments [equation]
        -> Context validity
        -> Typing judgements
        -> Equality judgements
\end{wstructure}

In the following, we adopt a standard presentation for our type
theory. It is defined by three forms of judgements: context validity
judgements, typing judgements, and equality judgements. Formally,
these judgments are presented as follow:

\[
\begin{array}{ll}
\Gamma \vdash \Valid                & \mbox{\(\Gamma\) is a valid context, giving types to variables} \\
\Gamma \vdash \Bhab{t}{T}           & \mbox{term \(t\) has type \(T\) in context \(\Gamma\)} \\
\Gamma \vdash \Bhab{s \equiv t}{T}  & \mbox{\(s\) and \(t\) are equal at type \(T\) in context \(\Gamma\)} \\
\end{array}
\]

\begin{wstructure}
    <- Invariants [equation]
        -> By induction on derivations
\end{wstructure}

Each of these judgements is populated by a respective system of
inferences rules. These rules are formulated so as to ensure that the
following implications always hold by induction on derivations:

\[
\begin{array}{l@{\;\;\Rightarrow\;\;}l}
\Gamma            \vdash \Bhab{t}{T}            
    & \Gamma \vdash \Valid\; \wedge\; \Gamma\vdash\Type{T} \\
\Gamma            \vdash s \equiv \Bhab{t}{T}   
    & \Gamma \vdash \Bhab{s}{T} \;\wedge\; \Gamma\vdash \Bhab{t}{T} \\
\Gamma ; \xS ; \Delta \vdash J                      
    & \Gamma \vdash \Bhab{s}{S} \;\Rightarrow\; 
          \Gamma ; \Delta[s/x] \vdash J[s/x] \\
\end{array}
\]

We are not going to prove the validity of these invariants. They
follow rather straightforwardly from the induction rules. For formal
proofs, we refer the reader to standard presentations of type theory,
such as Luo's seminal work \cite{luo:utt}.

\begin{wstructure}
    <- Judgemental equality
        <- Presentation independant of particular implementation choice
        -> Model in Agda, intensional
        -> Used in Epigram, OTT
\end{wstructure}

Noticeably, we have adopted a judgemental presentation of
equality. Therefore, we are not tied to a particular implementation
choice. In particular, our system has been modeled in Agda, which
features an intensional equality. On the other hand, it is used in
Epigram, which equality has a slightly extensional
flavor~\cite{altenkirch:ott}. We expect users of propositional or even
fully extensional systems to also find their way through this
presentation.

\begin{wstructure}
<- Context validity [figure]
    <- Not much to be said
\end{wstructure}

@@

The rules ensuring Context Validity (Fig.\ref{fig:context-validity})
are unsurprisingly common. They simply garantee the validity of the
empty context, as well as the linear extension of a context by a
well-typed variable.

\begin{wstructure}
<- Typing judgements [figure]
    <- Set in Set
        -> For simplicity of presentation
        -> Assume that a valid stratification can be inferred
            <- Harper-Pollack, Luo, Courant
        -> See later discussion
    <- Standard presentation of Pi and Sigma types
\end{wstructure}

The typing judgements (Fig.\ref{fig:typing-judgements}) are just as
familiar. We recognize the standard presentation of $\Pi$ and $\Sigma$
types, respectively inhabited by lambda terms and dependent
pairs. Naturally, there are rules for function application and
projections of $\Sigma$-types. Equal types can be substituted, thanks
to the conversion rule.

For the sake of presentation, we postulate a $\Set$ in $\Set$
rule. Having this rule makes our type theory inconsistent, by Girard's
paradox~\cite{girard:set-in-set}. However, it has been
shown~\cite{harper:implicit-universe, luo:utt,
  courant:explicit-universe} that a valid stratification can be
inferred, automatically or semi-automatically. In the remaining of our
presentation, we will assume that such a stratification exists, even
though we will keep it implicit. We shall discuss this assumption in
Section~\ref{sec:discussion}.

@@

\begin{wstructure}
<- Judgemental equality [figure]
    <- symmetry, reflexivity, and transitivity
    <- beta-rules for lambda and pair
    <- xi-rule for functions
    -> Agnostic in the notion of equality
        <- Doesn't rely on a ``propositional'' equality
        -> Key: wide applicability of our proposal
\end{wstructure}

Finally, we define the rules governing judgemental equality in
Figure~\ref{fig:judgemental-equality}. As one would expect,
judgemental equality is respects symmety, reflexivity, and
transitivity. Further, it also encompasses the $β$-rules for function
application and pairs. Finally, equality under lambda is dealt by the
$ξ$-rule.

Crucially, being judgemental, this presentation is agnostic in the
notion of equality actually implemented. Indeed, our typing and
equality judgements do not rely on a ``propositional'' equality. This
freedom is a key point in favour of the wide applicability of our
proposal. Obviously, when implementing our technology in an existing
type-theory, some opportunities arise. We will present some of them
along the course of the paper, clearly specifying which extensions we
require.

@@

\begin{wstructure}
!!! Need Help !!!
<- Meta-theoretical properties
    <- Assuming a stratified discipline
    <> The point here is to reassert that dependent types are not evil, 
       there is no non-terminating type-checker, or such horrible lies <>
    -> Strongly normalizing
        -> Every program terminates
    -> Type-checking terminates
    ???
\end{wstructure}

This completes our presentation of the type theory. Assuming a
stratified discpline of universe, the system we have described enjoy
some very strong meta-theoretical properties. Unlike simply typed
languages, such as Haskell, dependently-typed systems are
\emph{strongly normalizing}: every program that type-checks
terminates. Moreover, type-checking is decidable and can therefore be
implemented by a terminating algorithm.

@@

\begin{wstructure}
<- Motivation
    <- Finite sets could be encoded with Unit and Bool
        /> Hinder the ability to name things
    <- W-types considered harmful?
        ???
    -> For convenience
        <- Named elements
        <- Referring by name instead of code
        -> Types as coding presentation
            /> Also as coding representation!
\end{wstructure}

As a motivating example, we are now going to extend the type theory
with a notion of finite set. One could argue that there is no need for
such an extension: finite sets, just as any data-structure, can be
encoded inside the type theory. A well-known example of such encoding
is the Church encoding of natural numbers, which is isomorphic to
finite sets.

% Another example is Martin-Löf's W-types~\cite{}.

However, using encodings is impractical. In the case of finite sets,
for instance, we would like to name the elements of the sets. Then, we
need to be able to manipulate these elements by their name, instead of
their encoding. While we are able to give names to encodings, it is
extremely tedious to map the encodings back to a name. Whereas these
objects have a structure, the structure is lost during the encoding,
when they become anonymous inhabitants of a $Π$ or $Σ$-type.

In the simply-typed world, we are used to see types as a coding
presentation -- presentation of invariants, presentation of
properties. In the dependently-typed world, we also learn to use types
as a coding representation: finite sets being good citizens, they
ought to be democratically represented at the type level. As we will
see, this gives us the ability to name and manipulate them (this is
were the democracy analogy goes crazy, I think).

\begin{wstructure}
<- Implementation [figure]      
    <- Tags
        -> Purely informational token
    <- EnumU
        -> Universe of finite sets
    <- EnumT e
        -> Elements of finite set e
\end{wstructure}

The implementation of finite sets is presented in
Figure~\ref{fig:typing-finite-set}. It is composed by three
components. First, we define tags as inhabitants of the $\UId$ type. A
tag is solely an informative token, used for display purposes. Finite
sets inhabits the $\EnumU$ type. Unfolding the definition, we get that
a finite set is a list of tags. Finally, elements of a finite set $u$
belong to the corresponding $\EnumT{u}$ type. Intuitively, it
corresponds to an index, a number, pointing to an element of $u$.

@@

\begin{wstructure}
<- Equipment
    <- \spi operator
        <- Equivalent of Pi on finite sets
        <- First argument: (finite) domain
        <- Second argument: for each element of the domain, a co-domain
        -> Inhabitant of \spi: right-nested tuple of solutions
            <- Skip code for space reasons
    <- switch operator
        <- case analyses over x
        <- index into the \spi tuple to retrieve the corresponding result
\end{wstructure}

The $\Enum$ type comes with further machinery. Indeed, to manipulate
the $\EnumU$ and $\EnumT{}$ objects, we need to equip our type theory
with an appropriate \emph{elimination principle}. A first step
consists in defining the $\spi{}{}$ operator:

\[
%% spi
\spi{}{} : \PITEL{\V{e}}{\EnumU}
           \PITEL{\V{P}}{\EnumT{e} \To \Set} \To \Set 
\]

A closer look to its type should reveal a certain familiarity with its
equivalent for sets, $\Pi$. Indeed, $\spi{}{}$ defines a \emph{finite
  function space}, mapping elements of finite sets to objects in
$\Set$. Intuitively, this operator build a right-nested tuple,
effectively mapping an object of $P\:i$ for each $i$ the finite
domain.

Having understood the meaning of $\spi{}{}$, the definition of the
$\F{switch}$ eliminator is not surprising:

\[
%% switch
\begin{array}{@{}ll}
\F{switch} : & \PITEL{\V{e}}{\EnumU}
               \PITEL{\V{P}}{\EnumT{e} \To \Set} \\
             & \PITEL{\V{b}}{\spi{e}{P}}
               \PITEL{\V{x}}{\EnumT{e}} \To P~x
\end{array}
\]

Again, there is a clear equivalent in the full-$\Set$ world: function
application. The operational behavior of $\F{switch}$ is
straightforward: $x$ is peeled off as we move deeper inside the nested
tuple $b$. When $x$ equals $\Ze$, we simply return the value we are
pointing to.

\begin{wstructure}
<- Equivalent to having a function space over finite sets
    /> Made non-obvious by low-level encodings
        <- General issue with codes
             -> Need to provide an attractive presentation to the user
    -> Types seem to obfuscate our reading
        <- Provide ``too much'' information
        /> False impression: information is actually waiting to be used more widely
        -> See next Section
\end{wstructure}

As we have seen, finite sets are not just a way of making
data. Through the $\spi{}{}$ and $\F{switch}$ operators, they also
come with a notion of finite function space. However, we had to
extract that intuition from the type, by a careful reading. At a first
glance, this seems to contradict our argument in favor of types for
coding representation. Here, we are overflown by low-level, very
precise type information.

However, our situation is significantly different from the one we
faced with encoded data: while we were suffering from a crucial lack
of information, we are here facing too much information, hence loosing
focus. This is a general issue with the usage of codes, as they convey
much more information than what the user is used to see.

As we will see in the following section, there exists a cure to this
problem. In a nutshell, instead of being overflown by typing
information, we will put it at work, automatically. The consequence is
that, in such system, working with codes is \emph{practical}: one
should not be worried by information overload, but how to use it as
much as possible. Having mastered this difficulty, we should not be
afraid by code for practical purposes.

@@

\begin{wstructure}
<- Bidirectional type-checking [ref. Turner,Pierce]
    -> Separating type-checking from type synthesis
    <- Type checking: push terms into types
        <- Example: |Pi S T :>: \ x . t| allows us to drop annotation on lambda
    <- Type synthesis: pull types out of terms
        <- Example: |x : S l- x :<: S| gives us the type of x
\end{wstructure}

Our proposal to tame the problem of codes is deeply rooted in the
bidirectional presentation of type-checking of Pierce and
Turner~\cite{pierce:bidirectional-tc}. In this formalism,
type-checking is divided into two complementary phases. In the
\emph{type inference} phase, types are \emph{pulled} out of terms. A
typical example is a variable in the context:

\[
\Axiom{\Gamma ; \xS ; \Delta \Vdash \propag{\x}
                                           {\pull{\x}{S}}}
\]

Because the variable and its type are stored in the context, whenever
this variable is used, we can pull its type out of it.

On the other hand, in the \emph{type checking} phase, terms are
\emph{pushed} into types. We are handed a type together with a term,
our task consists in checking that the term is indeed of this type. We
are therefore allowed, and actually encouraged, to use the information
provided by the type. Therefore, we can relax our requirements on the
term. A typical example is the lambda:

\[
\Rule{\Gamma ; \xS \Vdash \propag{\push{t}{T\: x}}
                                 {t'}}
     {\Gamma \Vdash \propag{\push{\LAM{\x} t}{\PiTy{S}{T}}}
                           {\PLAM{\x}{S} t'}} 
\]

In our formal presentation of the type theory, we were forced to
annotate the lambda by the type of its domain. In a bidirectional
setting, the domain is provided by the $\Pi$ type we are pushing the
lambda into. Hence we do not need to annotate the lambda.

\begin{wstructure}
<- Formalization: type propagation
    <- Motivation
        -> High-level syntax
            -> exprIn: types are pushed in
                <- Subject to type *checking*
            -> exprEx: types are pulled from
                <- Subject to type *synthesis*
        -> Translated into our low-level type theory
        -> Presented as judgements
    -> Presentation mirors typing rule of [figure] 
        -> Ignore identitical judgements
\end{wstructure}

We adapt and formalize this \emph{type propagation} mechanism to our
richer type theory. First of all, we define a high-level syntax,
composed by \emph{expressions}. Expressions divided into two syntactic
categories: $\exprIn$ into which types are pushed in, and $\exprEx$
from which types are pulled out. Following the bidirectional
presentation, $\exprIn$ are subject to type \emph{checking}, while
$\exprEx$ are subject to type \emph{synthesis}.

The purpose of type propagation is to \emph{elaborate} an high-level
expression into a term of the type theory, together with its
type. This presentation builds on the experience gained in
implementing Epigram 1~\cite{mcbride:view-from-the-left}. Being
entirely syntax driven, we present our formalism as a system of
syntax-directed inference rules. As the presentation largely mirrors
the inference rules of the type theory, we will ignore the judgements
which are identical. We refer our reader to the associated technical
report~\cite{chapman:desc-tech-report} for the complete system of
rules.

\begin{wstructure}
<- Type synthesis [figure]
    <- Pull a type out of an exprEx
    <- Result in a full term, together with its type
    -> Do *not* need to specify types
        -> Extracting a term from the context
        -> Function application
        -> Projections
\end{wstructure}

The type synthesis rules are presented
Figure~\ref{fig:type-synthesis}. As illustrated by the judgement form,
a type synthesis problem is provided an expression $\exprEx$ in a
context, which is elaborated into a low-level term together with its
type. These terms act as \emph{sources} for the type propagation
machinery: they make types out of expressions. We have already studied
the case of a variable in context. This also include function
application as well as projections. Further down the line, these types
might propagate to other terms, thanks to type checking.

\begin{figure}
\input{figure_type_synthesis}
\caption{Type synthesis}
\label{fig:type-synthesis}
\end{figure}

\begin{wstructure}
<- Type checking [figure]
    <- Push a type in an exprIn
    <- Result in a full term
    -> *Use* the type to build the term!
        -> Domain and co-domain propagation for Pi and Sigma
        -> Translation of 'tags into EnumTs
        -> Translation of ['tags ...] into EnumUs
        -> Finite function space into switch
\end{wstructure}

Dually to type synthesis, type checking judgements
(Fig.\ref{fig:type-checking}) act as \emph{sinks}. From an expression
$\exprIn$ and a type propagated down to it, they elaborate a low-level
term. This is precisely the availability of the type we are pushed
into that makes the power of this technique. Its benefit on lambda
abstraction and pairs is well-known. Let us explore the consequences
on our coding scheme for finite sets.

A first consequence is that we can actually denote elements of finite
sets \emph{by their name}. Indeed, such expression will be pushed into
a $\EnumT{u}$, where $\Bhab{u}{\EnumU}$ contains all the names
populating the finite set. It is therefore easy to elaborate a name
into the corresponding index.

A second consequence concerns the definition of a finite set. We can
abstract away the details of code and use a familiar syntax
instead. Hence, we use the syntax $\Enum\: [t_1 \ldots t_n]$ to denote
the finite sets composed by the nammed elements $t_1$, \ldots,
$t_n$. The translation to elements of $\EnumU$ is straightforward.

Finally, and more importantly, we are able to present our user an
abstraction of \emph{finite} function space. This work as follow. When
an expression $t$ is pushed into a function which domain is a finite
set, we instead elaborate it as a term in a small $\pi$-type. That is,
a finite function. To realize our promise of a big $\Pi$-type, we
elaborate a call to the $\F{switch}$ eliminator.

\begin{figure}
\input{figure_type_checking}
\caption{Type checking}
\label{fig:type-checking}
\end{figure}
